TEMPERATURES AND METALLICITIES FOR 10,000 M DWARFS\\ IN THE \apogee\ SURVEY

### 2018-08-16 paper/draft.tex
add to caption
open questions

\begin{abstract}
\thecannon\ is a flexible, data-driven spectral modeling and parameter inference framework, demonstrated on high-resolution, near-infrared \apogee\ spectra of giant stars to estimate stellar labels (\teff, logg, [Fe/H], and detailed abundances) to precisions higher than the \apogee\ survey model-grid pipelines.
The lack of reliable stellar models in the near infrared for temperatures $<3500$\,K motivates the extension of this approach to M-dwarf stars, but with new training labels from optical surveys that overlap \sdssiv\ \apogee.
Using a training sample of 87 M dwarfs in the temperature range $2859 < T_{\rm eff} < 4131$\,K, and metallicty range $-0.48 < \mathrm{[Fe/H]}< 0.49$\,dex, we train a two-parameter model with predictive accuracy (in cross-validation) to 77\,K and 0.09\,dex respectively.
These are parameter-space precision statements, but the model is accurate in the data space: It accurately predicts \apogee\ spectral pixels at the percent level.
We also train a one-dimensional spectral type model using 51 M dwarfs with known spectral types ranging from M0 to M9, and demonstrate that \thecannon\ acting on \apogee\ spectra can infer spectral types with predictive accuracy of 0.9 types.
The trained models are employed to determine spectral subtypes, temperatures and metallicities for 10,311 M dwarfs in \apogee, selected in part on \gaia\ photometry and astrometry.
We validate our results using photometric temperature indicators and the \gaia\ color--magnitude diagram.
As a side result, we determine empirical relationships between spectral type and temperature and metallicity.
These results have implications for the NASA \tess\ Mission and other projects that involve large samples of M dwarfs.
\end{abstract}

@@

A number of M dwarf studies have already been conducted in \apogee\, particularly towards measuring reliable fundamental atmospheric parameters and kinematic measurements:
\begin{enumerate}
\item \citealt{Schmidt:2016} -- analyzed trends and reliablility of ASPCAP parameters for K/M dwarfs
\item \citealt{Souto:2017} and \citealt{Souto:2018} -- modeled three exoplanet hosting M dwarfs (Kepler-138, Kepler-168 and Ross-128), determining Teff/logg/metallicity + 13 elemental abundances
\item \citealt{Desphande:2013} -- modeled rotational and radial velocities for 200+ M dwarfs, later extended by \citealt{Gilhool:2018} to 700+ M dwarfs
\item \citealt{Rajpurohit:2018} -- tested BTSettl model grids on 45 M dwarfs, estimated Teff/logg/metallicity)
\item \citealt{Skinner:2018} - identified and measured mass ratios and radial velocities for 44 M dwarf binaries
\end{enumerate}

@@

The data-driven approach of \thecannon\ makes sense in various contexts:
One is the context in which stellar labels are only known for a small number of stars
but there are spectra taken for many more.
Another is the context in which it is very expensive to obtain labels for a star, but
there are many stars that need labels.
Another is the context in which there are spectral models that work in one wavelength
range or resolution but not in another.
The context of this paper is the latter:
There are working and reliable models of M dwarf stars in the optical, which have been
used successfully, but (as we note above) there isn't as much success with such modeling in the infrared,
and not at \apogee\ resolution.

@@

\begin{figure}[ht]
% \gridline{\fig{figures/safe_selection_teff.png}{0.5\textwidth}{}
%           \fig{figures/safe_selection_feh.png}{0.5\textwidth}{}}
% \gridline{\fig{figures/safe_selection_spt.png}{0.5\textwidth}{}}
\plottwo{figures/cmd_teff_full.png}{figures/cmd_teff_safe.png}
\plottwo{figures/cmd_feh_full.png}{figures/cmd_feh_safe.png}
\plottwo{figures/cmd_spt_full.png}{figures/cmd_spt_safe.png}
\caption{Full sample of 14,827 M dwarfs colored by \cannon\ labels before selection (left) and final sample selection of 10,311 M dwarfs after applying selection criteria described in Section \ref{subsec:test_selection} (right), to reduce contamination from sources with that are not similar to the training sample (not single, main-sequence M stars, such as pre-main sequence, binaries, and K dwarfs). Temperature gradient increases with decreasing color, spectral subtype increases with increasing color, and metallicity gradient increases with increasing absolute magnitude as expected. Deviations from these gradients seen at the upper slice of the main sequence is likely remaining contamination from the binary sequence. \label{fig:safe_selection}}
\end{figure}

>>>>

\begin{abstract}
\thecannon\ is a flexible, data-driven spectral modeling and parameter inference framework, demonstrated on high-resolution, near-infrared \apogee\ spectra of giant stars to estimate stellar labels (\teff, logg, [Fe/H], and detailed abundances) to precisions higher than the \apogee\ survey model-grid pipelines.
The lack of reliable stellar models in the near infrared for temperatures $<3500$\,K motivates the extension of this approach to M-dwarf stars, but with new training labels from optical surveys that overlap \sdssiv\ \apogee.
Using a training sample of 87 M dwarfs in the temperature range $2859 < T_{\rm eff} < 4131$\,K, and metallicty range $-0.48 < \mathrm{[Fe/H]}< 0.49$\,dex, we train a two-parameter model with predictive accuracy (in cross-validation) to 77\,K and 0.09\,dex respectively.
These are parameter-space precision statements, but the model is accurate in the data space: It accurately predicts \apogee\ spectral pixels at the percent level.
We also train a one-dimensional spectral type model using 51 M dwarfs with known spectral types ranging from M0 to M9, and demonstrate that \thecannon\ acting on \apogee\ spectra can infer spectral types with predictive accuracy of 0.9 types.
The trained models are employed to determine spectral subtypes, temperatures and metallicities for 10,311 M dwarfs in \apogee, selected in part on \gaia\ photometry and astrometry.
We validate our results using photometric temperature indicators and the \gaia\ color--magnitude diagram.
As a side result, we determine empirical relationships between spectral type and temperature and metallicity.
These results have implications for the NASA \tess\ Mission and other projects that involve large samples of M dwarfs.

\color{gcolor}{HOGG: just a reflection, I think the scope of this paper has shifted a lot from just being a technical paper about \thecannon\ working on M dwarfs, to drifting more towards M dwarf science (partly because I've learned a lot more about the properties/characteristics of M dwarfs, and the Gaia has given us a lot of intuition that we didn't have before). Anyways we should decide whether the paper is more of the former or the latter and maybe re-write the abstract. Right now it just feels a bit hacked-together, since this is basically just my AAS abstract that we kept adding some sentences to as we went, and I think it needs a better frame. I'm still trying to get a sense of what is scientifically interesting about this sample, and I've been struggling to pin point what the real relevance of this work is: it's not exactly just that the Cannon is precise for M dwarfs (this original scope, like an extension of Ness, is too boring I think, given what we've learned), and it's not really about M dwarf physics or why we are better than models (this is too beyond us to argue).

I think maybe the biggest angle is that are illustrating what can be leveraged out of two very high quality datasets (\apogee\ and \gaia), and we are the first to identify and collect a large (atmospheric parameters, 6D phase space) that have in the past been very difficult to measure. Is this what we want to emphasize?

Also an open question along these lines is whether we want to estimate any additional physical parameters for our sample, for example maybe adding radii (from the \citealt{Mann:2015} \teff-radius or $M_K$-radius relations), or masses from his $M_K$-mass relation he's publishing, or maybe there are some other fundamental parameters that would be interesting (and straightforward) to add? It seems like just because we have \gaia\ parallaxes we should go ahead and do it? Maybe also to do a full-circle comparison with our test sample labels.

Another open question is about the [Fe/H] validation, and whether we are happy with just the cmd plot of metallicity. I think our ideas for other tests were:
\begin{itemize}
\item Looking at M+M binary pairs: there are 174 pairs and it looks by eye that most of the pair lines are parallel to the main sequence. I don't know if there's some more clever way to see if the pairs lie in the same metallicity space of the cmd, so this is probably not that useful
\item Overplotting some isochrones at varying metallicities--which I'm still working on
\end{itemize}
}\color{black}

Also I still have to make the vertical action plot.
\end{abstract}

@@

A number of M dwarf studies have already been conducted in \apogee, particularly towards measuring reliable fundamental atmospheric parameters and kinematic measurements:
\begin{enumerate}
\item \citealt{Schmidt:2016} -- analyzed trends and reliablility of ASPCAP parameters for K/M dwarfs
\item \citealt{Souto:2017} and \citealt{Souto:2018} -- modeled three exoplanet hosting M dwarfs (Kepler-138, Kepler-168 and Ross-128), determining Teff/logg/metallicity + 13 elemental abundances
\item \citealt{Desphande:2013} -- modeled rotational and radial velocities for 200+ M dwarfs, later extended by \citealt{Gilhool:2018} to 700+ M dwarfs
\item \citealt{Rajpurohit:2018} -- tested BTSettl model grids on 45 M dwarfs, estimated Teff/logg/metallicity)
\item \citealt{Skinner:2018} - identified and measured mass ratios and radial velocities for 44 M dwarf binaries
\end{enumerate}

@@

The data-driven approach of \thecannon\ makes sense in various contexts:
One is the context in which stellar labels are only known for a small number of stars
but there are spectra taken for many more.
Another is the context in which it is very expensive to obtain labels for a star, but
there are many stars that need labels.
Another is the context in which there are spectral models that work in one wavelength
range or resolution but not in another.
The context of this paper is the latter:
There are working and reliable models of M dwarf stars in the optical, which have been
used successfully, but (as we note above) there isn't as much success with such modeling in the infrared,
and not at \apogee\ resolution.
\color{gcolor}{HOGG: I think we can make a more specific set of points here that would be more interesting to M dwarf people: temperatures and (especially) metallicities are difficult to measure accurately for an M dwarf, but people have done a lot of work to build calibrated empirical relations for metallicities and temperatures that are very precise. 
For example \citealt{Terrien:2012}, \citealt{RojasAyala:2010}, Mann 2013 (and there's probably some other examples (Newton, etc.) and I have more literature to look at...) have a metallicity callibration based off of an FGK+M star samples, from J, H, K or optical band spectra using the equivalent width of some strong line/index (like CaI, KI, etc). However these metallicity callibrations are inaccesible for APOGEE spectra, because the wavelength coverage does not include the relevant callibration indices. \thecannon\ bridges this gap: it doesn't require using specific lines (which would probably be a mess to deconvolve at high resolution with other blends even if we had the right lines); it allows us to use a lot of low level metallicity information present in thousands of lines; and (as we demonstrate) it does so with very good precision.}\color{black}

@@

\begin{figure}[ht]
% \gridline{\fig{figures/safe_selection_teff.png}{0.5\textwidth}{}
%           \fig{figures/safe_selection_feh.png}{0.5\textwidth}{}}
% \gridline{\fig{figures/safe_selection_spt.png}{0.5\textwidth}{}}
\plottwo{figures/cmd_teff_full.png}{figures/cmd_teff_safe.png}
\plottwo{figures/cmd_feh_full.png}{figures/cmd_feh_safe.png}
\plottwo{figures/cmd_spt_full.png}{figures/cmd_spt_safe.png}
\caption{Full sample of 14,827 M dwarfs colored by \cannon\ labels before selection (left) and final sample selection of 10,311 M dwarfs after applying selection criteria described in Section \ref{subsec:test_selection} (right), to reduce contamination from sources with that are not similar to the training sample (not single, main-sequence M stars, such as pre-main sequence, binaries, and K dwarfs). Overplotted with orange triangles are the M15 and W11 training samples, for their respective \cannon\ test labels. Temperature gradient increases with decreasing color, spectral subtype increases with increasing color, and metallicity gradient increases with increasing absolute magnitude as expected. Deviations from these gradients seen at the upper slice of the main sequence is likely remaining contamination from the binary sequence. \label{fig:safe_selection}}
\end{figure}
